---
title: "Homework 2"
author: "Jenny Eberling"
date: "2/2/2022"
output: word_document
---

```{r setup, echo = T, results = 'hide'}
knitr::opts_chunk$set(echo = TRUE)
snps <- read.csv("C:/Users/Owner/OneDrive/Desktop/Semester IV/Genetics/snps.csv")

# Population 1
mouse.freqs <- c(0,0,0,0,0,1,1,1,1,1)


# Population 2
blue.flower.freqs <- c(0.573, 0.504, 0.717, 0.657, 0.302, 0.339,
                       0.032, 0.007, 0.008, 0.005, 0.009, 0, 0, 0, 0, 0.005, 0.010, 0, 0, 0.004, 0.126, 0.068, 0.002, 0, 0, 0,
                       0.106, 0.224, 0.411, 0.014)

# Population 3
all.same.freqs <- c(0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 
                    0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 
                    0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 0.453, 0.453)
```

## Part 1

```{r }
# Function to calculate Fst
calc.Fst <- function(af){
  # determine length of allele frequency vector
  n<-length(af)
  
  # calculate H_s
  # need to calculate expected heterozygosity for each subpopulation
  # which is 2pq for each given allele frequency
  H_s <- sum(2*af*(1-af))/n
  
  # calculate H_t
  p_hat <- sum(af)/n
  H_t <- 2*p_hat*(1-p_hat)
  
  # calculate F_st
  F_st <- (H_t - H_s)/H_t
  
  # print and return the answer
  print(paste0("your F_st is: ", F_st))
  #return(F_st) # include if you need this number
}

# the Fst stat for each given population
calc.Fst(mouse.freqs)
calc.Fst(blue.flower.freqs)
calc.Fst(all.same.freqs)
```

Does it make sense?  

The F_st statistic compares the heterozygosity of subpopulations to the heterozygosity of the total population. 

* In the mouse example, the mice of AA and aa are kept separate. Thus there is 0 heterozygosity, so H_s is 0, making the F_st statistic H_t / H_t = 1. A F_st of 1 means that the heterozygosity in the subpopulation is super small, and in this example there is no possibility of heterozygosity, so an F_st of 1 is expected.  
  
* In the flower example, there are a couple different substructures, so we expect the F_st to be somewhere between 0 and 1. Also note there are quite a few subpopulations with 0 heterozygosity, so we expect the F_st to be closer to 0 than 1. The resulting .399 fulfills this.  
  
* In the same frequency example, every subpopulation has the same heterozygosity as the total population, so H_s munus H_t will result in a 0 numerator. 0 means there is 'little to no population substructure,' and in this example there are no substructures (the subpop are the same as the large pop), so 0 is what we'd expect. 



## Part 2

```{r, echo = T, results = 'hide'}
s1<-ifelse(snps$snp1 == 'CC',1,ifelse(snps$snp1 == 'CT',2,3)) # CC=1, CT=2, TT=3
s2<-ifelse(snps$snp2 == 'CC',1,ifelse(snps$snp2 == 'CT',2,3))
s3<-ifelse(snps$snp3 == 'GG',1,ifelse(snps$snp3 == 'TG',2,3))
s4<-ifelse(snps$snp4 == 'AA',1,ifelse(snps$snp4 == 'AT',2,3))
s5<-ifelse(snps$snp5 == 'CC',1,ifelse(snps$snp5 == 'CG',2,3))
s6<-ifelse(snps$snp6 == 'CC',1,ifelse(snps$snp6 == 'CG',2,3))
s7<-ifelse(snps$snp7 == 'CC',1,ifelse(snps$snp7 == 'GC',2,3))
s8<-ifelse(snps$snp8 == 'GG',1,ifelse(snps$snp8 == 'GT',2,3))
s9<-ifelse(snps$snp9 == 'AA',1,ifelse(snps$snp9 == 'AC',2,3))
s10<-ifelse(snps$snp10 == 'CC',1,ifelse(snps$snp10 == 'CT',2,3))
s11<-ifelse(snps$snp11 == 'CC',1,ifelse(snps$snp11 == 'CT',2,3))

snps.num <- cbind(s1,s2,s3,s4,s5,s6,s7,s8,s9,s10,s11)

```

Now we can do what is on 'simple example'  

```{r }
pca.output <- prcomp(snps.num) #phi values. use them in linear comb to get z (called x in R)
summary(pca.output)
head(snps.num)
#scatterplot of PC1 and PC2
x1 <- pca.output$x[,1]
x2 <- pca.output$x[,2]
plot(x1,x2)
```

PC and variance:  
According to 'cumulative proportion of variance' it takes 4 pc's to explain 50.9% of the variance and 10 to explain 95% percent of the variance (although with 9 you can explain 88% which is close to 90%).

Scatter plot comments:   
PC1 centers around 3 values - -.5, .5, and 1.5 - while PC2 ranges from -1 to 2. What the graph says to me is that x2 explains variation that x1 does not, because it holds different values than x1.

No substructure, so random looking.


